\SW: Crowd-sourcing the Discovery of Gravitational Lenses

### 2015-03-17 doc/sw-system.tex
Done intro #96

\sw is a web-based service that enables the discovery of strong gravitational
lenses in wide-field imaging surveys by large numbers of people. Carefully
produced color composite images are displayed to volunteers via a
classification interface which records their estimates of the positions of
candidate lensed features. Simulated lenses, and expert-classified non-lenses,
are inserted into the image stream at random intervals; this training set is
used to give the volunteers feedback on their performance, and to estimate a
dynamically-updated probability for any given image to contain a lens. Low
probability systems are retired from the site periodically, concentrating the
sample towards a set of candidates; this ``stage 1'' set is then re-classified
by the volunteers in a second  refinement stage. Analyzing the classification
of the training set, we predict that the first stage alone should yield a
sample that is C\% complete, while leading to the rejection of R\% of the
initial target sample. Having divided the 150 square degree CFHTLS imaging
survey into 430000 overlapping 70 by 70 arcminute tiles and displayed them on
the site, we were joined by 33000 volunteers who contributed X million image
classifications over the course of N months. The sample was reduced to 3500
stage 1 candidates; these were then refined to yield a sample of 1400
candidates rankable by their stage 2 probability. We expect this sample to be
X\% complete and Y\% pure at a threshold of 95\% classification probability.
We find that, on average, and given the assumptions we make in our analysis,
we need 9 classifications per image during the first stage, X in the second.
We estimate the mean information contributed per person to be X bits, over a
session lasting, on average, N classifications per volunteer, and present the
highly skewed distributions of these quantities. We comment on the scalability
of the \sw system to the wide field survey era, and its potential to operate
beyond its design as a supervised classification system.

@@

Strong gravitational lensing -- the formation of multiple, magnified images of
background objects due to the deflection of light by  massive foreground
objects -- is a very powerful astrophysical tool, enabling a wide range of
science projects. The image separations and distortions provide information
about the mass distribution in the lens \citep[e.g.][]{AugerEtal2010,
SonnenfeldEtal2012,SonnenfeldEtal2013}, including on sub-galactic scales
\citep[e.g.][]{Dalal+Kochanek2002,VegettiEtal2010,HezavehEtal2013}. Any strong
lens can provide magnification of a factor of 10 or more, providing a deeper,
higher resolution view of the distant universe through these ``cosmic
telescopes'' \citep[e.g.][]{StarkEtal2008,NewtonEtal2011}. Lensed quasars
enable cosmography via the time delays between the multiple images'
lightcurves \citep[e.g.][]{TewesEtal2013,SuyuEtal2013}, and study of the
accretion disk itself through the microlensing effect
\citep[e.g.][]{PoindexterEtal2008}. All of these investigations would
benefit from being able to draw from a larger sample of lenses.

In the last decade the numbers of detections of these rare cosmic alignments
has increased by an order of magnitude, thanks to searches carried out in 
wide field surveys, such as
CLASS \citep[e.g.]{BrowneEtal2003}, SDSS \citep[e.g.][]
{BoltonEtal2006,AugerEtal2010b,TreuEtal2011,InadaEtal2012}, CFHTLS
\citep[e.g.][]{MoreEtal2012,GavazziEtal2014}, Herschel
\citep[][]{NegrelloEtal2014} and SPT \citep[e.g.][]{VieiraEtal2013}, among
others.  As the number of known lenses has increased, new types have been
discovered, leading to entirely new investigations. Compound lenses
\citep{GavazziEtal2008,CollettEtal2012} and lensed supernovae
\citep{QuimbyEtal2014,KellyEtal2014} are good examples of this.

@@

Because they are rare, strong lenses are expensive to find. The most efficient
searches to date have made use of relatively clean signals such as the
presence of emission or absorption features at two distinct redshifts in the
same optical spectrum \citep[e.g.][]{BoltonEtal2004}, or the strong
``magnification bias'' towards detecting strongly-lensed sources in the sub-mm
waveband \citep[e.g.][]{NegrelloEtal2010}. Such searches have to be efficient,
because they require expensive high resolution imaging follow-up; consequently
they have so far produced yields in the tens to hundreds. An alternative
approach is to search images already of sufficiently high resolution and color
contrast, and confirm the systems as gravitational lenses by modeling the
survey data themselves \citep[][]{MarshallEtal2009}. Several square degrees of
HST images have been searched, yielding several tens of galaxy-scale lenses
\citep[e.g.][]{MoustakasEtal2007,FaureEtal2008,Jackson2008,MoreEtal2012,
PawaseEtal2014}. Similarly, searches of over a hundred square degrees of CFHT
Legacy Survey (CFHTLS) ground-based imaging, also with sub-arcsecond image quality,
have revealed a smaller number of wider image separation group-scale systems
\citep[e.g.][]{CabanacEtal2007,MoreEtal2012}. Detecting galaxy-scale lenses
from the ground is hard, but feasible albeit with lower efficiency and requiring
HST or spectroscopic follow-up to confirm the candidates as lenses
\citep[e.g.][]{GavazziEtal2014}.

@@

How can we scale these lens searches up to imaging surveys covering a hundred
times the sky area, such as the almost-all sky surveys planned with LSST and
Euclid, while reducing our dependence on expensive follow-up confirmation
observations? There are two approaches to detecting lenses in imaging surveys.
The first one is robotic: automated analysis of the object catalogs and the
survey images. The candidate samples produced by these methods have, to date,
not been of high purity  \citep[see
e.g.][]{MarshallEtal2009,MoreEtal2012,GavazziEtal2014}, with visual inspection
by teams of humans still required to narrow down the robotically-generated
samples. In this approach, the image data may or may not be explicitly
modelled by the robots as if it contained a gravitational lens, but the visual
inspection can be thought of as a ``mental modeling'' step. Systems classified
by an inspector to be good lens candidates are deemed as such because the
features in the image can be explained by a model of what gravitational lenses
do contained in the inspector's brain. The second approach simply cuts out the
robot middleman: \citet{MoustakasEtal2007,FaureEtal2008,Jackson2008} and
\citet{PawaseEtal2014} all performed successful, entirely visual searches for
lenses in HST imaging.

Visual image inspection thus seems, at present, unavoidable at some level when
searching for gravitational lenses. The technique has some drawbacks, however.
First is that humans are only humans, and they make mistakes. A solution to
this is for the inspectors to operate in teams, providing multiple
classifications of the same images in order to catch errors and correct them.
Second, and relatedly, is that humans get tired. With a well-designed
classification interface, a human might be able to inspect images at a rate of
one astronomical object per second (provided the majority are indeed
uninteresting). At $10^4$ massive galaxies, and 10 lenses, per square degree,
visual lens searches in good quality imaging data are limited to a few square
degrees per inspector per day. Scaling to thousands of square degrees
therefore means either robotically reducing the number of targets for
inspection, or increasing the number of inspectors, or both.

@@

Systematic detection of rare astronomical objects by such ``crowd-sourced''
visual inspection has recently been achieved by the online citizen science
project PlanetHunters \citep{SchwambEtal2012}. PlanetHunters was designed to
enable the discovery of transiting exoplanets in data taken by the Kepler
satellite; a community of N inspectors from the general public found, after
each undergoing a small amount of training, N new exoplanet candidates by
visual inspection of the Kepler lightcurves that were presented in a custom
web-based classification interface. The older Galaxy Zoo morphological
classification project \citep{LintottEtal2008} has also enabled the discovery
of rare objects, via its flexible inspection interface and discussion forum
\citep{LintottEtal2009}. Indeed, several of us (AV,EB,CC,TJ,CM,LW) were active
in an informal Galaxy Zoo gravitational lens search, an experience which led
to the present hypothesis that a systematic online visual lens search could be
successful.

@@

In this paper, we describe the \sw web application, an online system that
enables  crowd-sourced gravitational lens detection by inviting volunteers to
classify astronomical survey images as containing lens candidates or not.  In
a companion paper (More et al, in preparation, herafter \papertwo) we will
present the new gravitational lenses discovered in our first, experimental,
lens search, and begin to investigate the differences between lens detections
made in \sw and those made with automated techniques. Here though, we simply
try to answer the following questions:

@@

Our basic method in this paper is to analyze the performance of the \sw system
on the ``training set'' of simulated lenses and known non-lenses. This allows
us to estimate completeness and purity with respect to gravitational lenses
that have the same properties of the training set. In \papertwo we carry out
similar tests but with a sample of known (real) lenses {\it in situ}.

>>>>

We describe \sw, a novel gravitaional lens discovery service that
yields samples of high purity and completeness through crowd-sourced
visual inspection.
Carefully
produced color composite images are displayed to volunteers via a
classification interface which records their estimates of the positions of
candidate lensed features. Simulated lenses, and expert-classified non-lenses,
are inserted into the image stream at random intervals; this training set is
used to give the volunteers feedback on their performance, and to estimate a
dynamically-updated probability for any given image to contain a lens. Low
probability systems are retired from the site periodically, concentrating the
sample towards a set of candidates; this ``stage 1'' set is then re-classified
by the volunteers in a second  refinement stage. Analyzing the classification
of the training set, we predict that the first stage alone should yield a
sample that is C\% complete, while leading to the rejection of R\% of the
initial target sample. Having divided the 150 square degree CFHTLS imaging
survey into 430000 overlapping 70 by 70 arcminute tiles and displayed them on
the site, we were joined by 33000 volunteers who contributed X million image
classifications over the course of N months. The sample was reduced to 3500
stage 1 candidates; these were then refined to yield a sample of 1400
candidates rankable by their stage 2 probability. We expect this sample to be
X\% complete and Y\% pure at a threshold of 95\% classification probability.
We find that, on average, and given the assumptions we make in our analysis,
we need 9 classifications per image during the first stage, X in the second.
We estimate the mean information contributed per person to be X bits, over a
session lasting, on average, N classifications per volunteer, and present the
highly skewed distributions of these quantities. We comment on the scalability
of the \sw system to the wide field survey era, and its potential to operate
beyond its design as a supervised classification system.

@@

Strong gravitational lensing -- the formation of multiple, magnified images of
background objects due to the deflection of light by  massive foreground
objects -- is a very powerful astrophysical tool, enabling a wide range of
science projects. The image separations and distortions provide information
about the mass distribution in the lens \citep[e.g.][]{AugerEtal2010,
SonnenfeldEtal2012,SonnenfeldEtal2013}, including on sub-galactic scales
\citep[e.g.][]{Dalal+Kochanek2002,VegettiEtal2010,HezavehEtal2013}. Any strong
lens can provide magnification of a factor of 10 or more, providing a deeper,
higher resolution view of the distant universe through these ``cosmic
telescopes'' \citep[e.g.][]{StarkEtal2008,NewtonEtal2011}. Lensed quasars
enable cosmography via the time delays between the multiple images'
lightcurves \citep[e.g.][]{TewesEtal2013,SuyuEtal2013}, and study of the
accretion disk itself through the microlensing effect
\citep[e.g.][]{PoindexterEtal2008}. All of these investigations would
benefit from being able to draw from a larger and/or more diverse sample of lenses.

In the last decade the number of these rare cosmic alignments known
has increased by an order of magnitude, thanks to searches carried out in
wide field surveys, such as
CLASS \citep[e.g.]{BrowneEtal2003}, SDSS \citep[e.g.][]
{BoltonEtal2006,AugerEtal2010b,TreuEtal2011,InadaEtal2012}, CFHTLS
\citep[e.g.][]{MoreEtal2012,GavazziEtal2014}, Herschel
\citep[][]{NegrelloEtal2014} and SPT \citep[e.g.][]{VieiraEtal2013}, among
others.  As the number of known lenses has increased, new types have been
discovered, leading to entirely new investigations. Compound lenses
\citep{GavazziEtal2008,CollettEtal2012} and lensed supernovae
\citep{QuimbyEtal2014,KellyEtal2014} are good examples of this.

@@

Because they are rare, strong lenses are expensive to find. The highest purity
searches to date have made use of relatively clean signals such as the
presence of emission or absorption features at two distinct redshifts in the
same optical spectrum \citep[e.g.][]{BoltonEtal2004}, or the strong
``magnification bias'' towards detecting strongly-lensed sources in the sub-mm
waveband \citep[e.g.][]{NegrelloEtal2010}. Such searches have to yield pure samples,
because they require expensive high resolution imaging follow-up; consequently
they have so far produced yields in the tens to hundreds. An alternative
approach is to search images already of sufficiently high resolution and color
contrast, and confirm the systems as gravitational lenses by modeling the
survey data themselves \citep[][]{MarshallEtal2009}. Several square degrees of
HST images have been searched, yielding several tens of galaxy-scale lenses
\citep[e.g.][]{MoustakasEtal2007,FaureEtal2008,Jackson2008,MoreEtal2012,
PawaseEtal2014}. Similarly, searches of over a hundred square degrees of CFHT
Legacy Survey (CFHTLS) ground-based imaging, also with sub-arcsecond image quality,
have revealed a smaller number of wider image separation group-scale systems
\citep[e.g.][]{CabanacEtal2007,MoreEtal2012}. Detecting galaxy-scale lenses
from the ground is hard, but feasible albeit with lower efficiency and requiring
HST or spectroscopic follow-up to confirm the candidates as lenses
\citep[e.g.][]{GavazziEtal2014}.

@@

How can we scale these lens searches up to imaging surveys covering a hundred
times the sky area, such as the almost-all sky surveys planned with LSST and
Euclid?  There are two approaches to detecting lenses in imaging surveys.
The first one is robotic: automated analysis of the object catalogs and the
survey images. The candidate samples produced by these methods have, to date,
not been of high purity  \citep[see
e.g.][]{MarshallEtal2009,MoreEtal2012,GavazziEtal2014}, with visual inspection
by teams of humans still required to narrow down the robotically-generated
samples. In this approach, the image data may or may not be explicitly
modelled by the robots as if it contained a gravitational lens, but the visual
inspection can be thought of as a ``mental modeling'' step. Systems classified
by an inspector to be good lens candidates are deemed as such because the
features in the image can be explained by a model of what gravitational lenses
do contained in the inspector's brain. The second approach simply cuts out the
robot middleman: \citet{MoustakasEtal2007,FaureEtal2008,Jackson2008} and
\citet{PawaseEtal2014} all performed successful, entirely visual searches for
lenses in HST imaging.

Visual image inspection thus seems, at present, unavoidable at some level when
searching for gravitational lenses. The technique has some drawbacks, however.
The first is that humans make mistakes. A solution to
this is for the inspectors to operate in teams, providing multiple
classifications of the same images in order to catch errors and correct them.
Second, and relatedly, is that humans get tired. With a well-designed
classification interface, a human might be able to inspect images at a rate of
one astronomical object per second (provided the majority are indeed
uninteresting). At $10^4$ massive galaxies, and 10 lenses, per square degree,
visual lens searches in good quality imaging data are limited to a few square
degrees per inspector per day (and less, if mre time is spent assessing difficult systems).
Scaling to thousands of square degrees
therefore means either robotically reducing the number of targets for
inspection, or increasing the number of inspectors, or both.

@@

Systematic detection of rare astronomical objects by such ``crowd-sourced''
visual inspection has recently been achieved by the online citizen science
project PlanetHunters \citep{SchwambEtal2012}. PlanetHunters was designed to
enable the discovery of transiting exoplanets in data taken by the Kepler
satellite; a community of N inspectors from the general public found, after
each undergoing a small amount of training, N new exoplanet candidates by
visual inspection of the Kepler lightcurves that were presented in a custom
web-based classification interface. The older Galaxy Zoo morphological
classification project \citep{LintottEtal2008} has also enabled the discovery
of rare objects, via its flexible inspection interface and discussion forum
\citep{LintottEtal2009}. Indeed, several of us (AV,CC,CM,EB,PM,TJ,LW) were active
in an informal Galaxy Zoo gravitational lens search (Verma et al, in preparation),
an experience which led
to the present hypothesis that a systematic online visual lens search could be
successful.

@@

In this paper, we describe the \sw web application, an
online system that was conceived to address the visual inspection problem in
gravitational lens detection for future large surveys via crowd-sourcing, and
implemented as a Zooniverse project.
In a companion paper (More et al, in preparation, herafter \papertwo) we will
present the new gravitational lenses discovered in our first, experimental,
lens search, and begin to investigate the differences between lens detections
made in \sw and those made with automated techniques. Here though, we simply
try to answer the following questions:

@@

Our basic method in this paper is to analyze the performance of the \sw system
on the ``training set'' of simulated lenses and known non-lenses. This allows
us to estimate completeness and purity with respect to gravitational lenses
that have the same properties of the training set. In \papertwo,
we carry out a complementary analysis using a sample of
``known'' (reported in the literature) lenses.
